---
title: "Ex5"
author: "Guanting Wei"
date: "Oct.7.2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Verifying E-step and M-step
\[
\begin{aligned}
f(y_i|x_i,\Psi)&=\sum_{j=1}^{m}\pi_j\phi(y_i;x_{i}^{T}\beta_j,\sigma^2 )\\&=\sum_{j=1}^{m}\pi_j\phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2 )
\end{aligned}
\]

\[
\begin{aligned}
l_{c}^{n}(\Psi )&=\sum_{i=1}^{n}\sum_{j=1}^{m}z_{ij}\ln\{\pi_j\phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2 )\}\\&=\sum_{i=1}^{n}\sum_{j=1}^{m}z_{ij}\{\ln \pi_j+\ln \phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2)\}
\end{aligned}
\]

\[
\begin{aligned}
E[z_{ij}|x_i,y_i;\Psi^{(k)}]&=\frac{\pi_{j}^{(k)}\phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2)}{\sum_{j=1}^{m}\pi_{j}^{(k)}\phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2)}\\&=p_{ij}^{(k+1)}
\end{aligned}
\]

\[
y=\begin{pmatrix}
y_{1}\\ 
y_{2}\\ 
...\\
y_i\\
...\\ 
y_{n}
\end{pmatrix}\
x=\begin{pmatrix}
x_1&x_2&...&x_i&...&x_n 
\end{pmatrix}^T=\begin{pmatrix}
x_{11} &x_{12}  &...  &x_{1p} \\ 
x_{21} &x_{22}  &...  &x_{2p} \\ 
... &...  &...  &... \\ 
x_{i1} &x_{i2}  &...  &x_{ip} \\ 
... &...  &...  &... \\ 
x_{n1} &x_{n2}  &...  &x_{np} 
\end{pmatrix}
\beta_j=\begin{pmatrix}
\beta_{j1}\\ 
\beta_{j2}\\ 
...\\ 
\beta_{jp}
\end{pmatrix}
\]


## 1.1 E-step
\[
\begin{aligned}
Q(\Psi |\Psi^{(k)})&=E[l_{n}^{c}(\Psi)|x,y,\Psi^{(k)}]\\&=\sum_{z}l_{n}^{c}(\Psi)P(z|x,y,\Psi^{(k)})
\\&=\sum_{z}P(z|x,y,\Psi^{(k)})\sum_{i=1}^{n}\sum_{j=1}^{m}z_{ij}\{\ln \pi_j+\ln \phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2)\}
\\&=\sum_{i=1}^{n}\sum_{j=1}^{m}[\sum_{z}P(z|x,y,\Psi^{(k)})z_{ij}]\{\ln \pi_j+\ln \phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2)\}\\&=\sum_{i=1}^{n}\sum_{j=1}^{m}E[z_{ij}|x_i,y_i;\Psi^{(k)}]\{\ln \pi_j+\ln \phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2)\}\\&=\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}\{\ln \pi_j+\ln \phi(y_i-x_{i}^{T}\beta_j;0,\sigma^2)\}\\&=\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}\{\ln \pi_j-\frac{1}{2}\ln 2\pi\sigma^2-\frac{1}{2}\frac{(y_i-x_i^T\beta_j)^2}{\sigma^2}\}
\end{aligned}
\]

## 1.2 M-step
### 1.2.1 For $\pi_j$
\[\
\begin{aligned}
\frac{\partial Q}{\partial \pi_j}&=\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}\ln\pi_j\\&=\sum_{j=1}^{m}(\ln\pi_j\sum_{i=1}^{n}p_{ij}^{(k+1)})\\&=\ln\pi_1\sum_{i=1}^{n}p_{i1}^{(k+1)}+...+\ln\pi_j\sum_{i=1}^{n}p_{ij}^{(k+1)}+...\ln\pi_m\sum_{i=1}^{n}p_{im}^{(k+1)}\\&=\ln\pi_1\sum_{i=1}^{n}p_{i1}^{(k+1)}+...+\ln\pi_j\sum_{i=1}^{n}p_{ij}^{(k+1)}+...\ln(1-\pi_1-...-\pi_j-...-\pi_{m-1})\sum_{i=1}^{n}p_{im}^{(k+1)}\\&=\frac{\sum_{i=1}^{n}p_{ij}}{\pi_j}-\frac{\sum_{i=1}^{n}p_{im}}{\pi_m}=0\\
\\
&\frac{\sum_{i=1}^{n}p_{ij}}{\pi_j}=\frac{\sum_{i=1}^{n}p_{im}}{\pi_m}\\\Rightarrow&\sum_{i=1}^{n}p_{ij}=\pi_j\frac{\sum_{i=1}^{n}p_{im}}{\pi_m}\\\Rightarrow&\sum_{j=1}^{m}\sum_{i=1}^{n}p_{ij}=\sum_{j=1}^{m}\pi_j\frac{\sum_{i=1}^{n}p_{im}}{\pi_m}\\\Rightarrow&\sum_{i=1}^{n}(\sum_{j=1}^{m}p_{ij})=\frac{\sum_{i=1}^{n}p_{im}}{\pi_m}(\sum_{j=1}^{m}\pi_j)\\\Rightarrow&\sum_{i=1}^{n}1=\frac{\sum_{i=1}^{n}p_{im}}{\pi_m}=n\\\Rightarrow&\frac{\sum_{i=1}^{n}p_{ij}}{\pi_j}=n\\\Rightarrow&\pi_j=\frac{\sum_{i=1}^{n}p_{ij}}{n}
\end{aligned}
\]

### 1.2.2 For $\beta_j$
\[
\begin{aligned}
\frac{\partial Q}{\partial\beta_j}&=-\frac{1}{2\sigma^2}\sum_{i=1}^{n}p_{ij}^{(k+1)}\frac{\partial (y_i-x_i^T\beta_j^{(k+1)})^2}{\partial \beta_j}\\&=-\frac{1}{2\sigma^2}\sum_{i=1}^{n}p_{ij}^{(k+1)}2(y_i-x_i^T\beta_j^{(k+1)})\frac{\partial (y_i-x_i^T\beta_j^{(k+1)})}{\partial \beta_j}\\&=-\frac{1}{2\sigma^2}\sum_{i=1}^{n}p_{ij}^{(k+1)}(-2)(y_i-x_i^T\beta_j^{(k+1)})\frac{\partial (x_i^T\beta_j^{(k+1)})}{\partial \beta_j}\\&=\frac{1}{\sigma^2}\sum_{i=1}^{n}p_{ij}^{(k+1)}x_i(y_i-x_i^T\beta_j^{(k+1)})=0\ (\beta_j\ is\ a\ column\ vector; we\ will\ get\ x_i^T\ if\ \beta_j\ is\ row\ vector)\\\Rightarrow&\beta_j^{(k+1)}=(\sum_{i=1}^{n}x_ix_i^Tp_{ij}^{(k+1)})^{-1}(\sum_{i=1}^{n}x_ip_{ij}^{(k+1)}y_i)
\end{aligned}
\]

### 1.2.3 For $\sigma^2$
\[
\begin{aligned}
&\frac{\partial Q}{\partial \sigma^2}=-\frac{1}{2\sigma^2}\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}+\frac{1}{2(\sigma^2)^2}\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}(y_i-x_i^T\beta_j^{(k+1)})^2=0\\\Rightarrow&\frac{\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}(y_i-x_i^T\beta_j^{(k+1)})^2}{\sigma^2}=n\\\Rightarrow&{\sigma^2}^{(k+1)}=\frac{\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}(y_i-x_i^T\beta_j^{(k+1)})^2}{n}
\end{aligned}
\]

# 2. EM function
```{r}
regmix_em=function(y,xmat,pi.init,beta.init,sigma.init,control=list(maxit=500,tol=1e-5)){
  n=nrow(xmat)
  p=ncol(xmat)
  m=length(pi.init)
  
  x=as.matrix(xmat)
  pi=pi.init
  beta=beta.init
  sigma=sigma.init
  
  maxit=control$maxit
  tol=control$tol
  
  P=matrix(0,n,m)
  beta.new=matrix(0,p,m)
  count=0
  
  for(iter in 1:maxit){
    for(i in 1:n){
      P[i, ]=pi*dnorm(y[i]-x[i, ]%*%beta,0,sigma)/sum(pi*dnorm(y[i]-x[i, ]%*%beta,0,sigma))
    }
    
    pi.new=colMeans(P)
    
    for(j in 1:m){
      beta.new[ ,j]<-solve(t(x)%*%diag(P[,j])%*%x)%*%t(x)%*%diag(P[,j])%*%y
    }
    
    sigma.new=sqrt(sum((P*(y%*%t(rep(1,m))-x%*%beta.new)^2))/n)
    
    if(sum(abs(pi.new-pi))+sum(abs(beta.new-beta))+sum(abs(sigma.new-sigma))<tol)break
      pi=pi.new
    beta=beta.new
    sigma=sigma.new
  }
  if(iter==maxit)
  print("reach maximun loop")
  list(pi=pi.new,beta=beta.new,sigma=sigma.new,iteration=iter)
}
```

# 3. Generate data
```{r}
regmix_sim <- function(n, pi, beta, sigma) {
  K <- ncol(beta)
  p <- nrow(beta)
  xmat <- matrix(rnorm(n * p), n, p) # normal covaraites
  error <- matrix(rnorm(n * K, sd = sigma), n, K)
  ymat <- xmat %*% beta + error # n by K matrix
  ind <- t(rmultinom(n, size = 1, prob = pi))
  y <- rowSums(ymat * ind)
  data.frame(y, xmat)
}
n <- 400
pi <- c(.3, .4, .3)
bet <- matrix(c( 1,  1,  1, 
                 -1, -1, -1), 2, 3)
sig <- 1
set.seed(1205)
dat <- regmix_sim(n, pi, bet, sig)
output1=regmix_em(y = dat[,1], xmat = dat[,-1], 
           pi.init = pi / pi / length(pi),
           beta.init = bet*0,
           sigma.init = sig / sig, 
           control = list(maxit = 500, tol = 1e-5))
output1
output2=regmix_em(y = dat[,1], xmat = dat[,-1], 
           pi.init = pi / pi / length(pi),
           beta.init = bet*1,
           sigma.init = sig / sig, 
           control = list(maxit = 500, tol = 1e-5))
output2
```

